Supermartingales and Termination Analysis of Statistical PCF

### 2021-10-16 LICS21/SPCF.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

\subsection{Trace semantics}
\label{sec:sampling semantics}
The relation $\to$ allows $\tsample$ to reduce to any number in $[0,1]$. 
This defines which results are possible, but to specify the exact probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \citep{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}. 
We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

>>>>

\subsection{Trace semantics}
\label{sec:sampling semantics}
The relation $\to$ allows $\tsample$ to reduce to any number in $[0,1]$. 
This defines which results are possible, but to specify the exact probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \defn{entropy space}, and elements of $\entrosp$ \defn{traces}.

Define a \defn{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \citep{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}. 
We say that a term $M$ is \defn{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

### 2021-10-16 LICS21/antitone.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

\begin{definition}
The definition of \emph{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
where $R \to R'$ is the $\tY$-redex rule.
Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\iffalse
Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
\[
f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
\fi
\end{definition}

@@

\begin{definition}
Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
we say that $(Y_n)_{n \geq 0}$ is an \emph{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\(
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
\).
%Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\end{definition}

@@

\begin{definition}
Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\end{definition}

\begin{therm}
%\begin{restatable}[Sparse function]{therm}{AntitonePartial}  
\label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
%\end{restatable}
\end{therm}

@@

Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

@@

\begin{example}[Higher-order recursion]
\label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$. Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function 
\[
F \, \underline{n_0} \, (F \, \underline{n_1} \, (\dots(\Xi\ F\, \underline x\, \underline n)\dots) \mapsto g_2(n),
\] 
where $g_2$ is as in Example~\ref{ex:unbiased random walk}, using the same antitone function too.

>>>>

\begin{definition}\rm
The definition of \defn{antitone ranking function} $f$ for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) \] 
where $R \to R'$ is the $\tY$-redex rule.
Any closed term for which an antitone ranking function exists is called \defn{antitone rankable}.
\iffalse
Given a ranking function $f$ on $M \in \Lambda^0$, we say that $f$ is \emph{antitone} if there exists an antitone function\footnote{i.e.~$r < r'$ implies $\epsilon(r) \geq \epsilon(r')$} $\epsilon : \nnReal \to \pReal$ such that %the ranking function $f$ satisfies
\[
f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
for all $E[R'] \in \mathit{Rch}(M)$ where $R \to R'$ is the $\tY$-redex rule.
\fi
\end{definition}

@@

\begin{definition}\rm
Given a probability space $(\Omega, \calF, \mathbb{P})$, and a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$,
we say that $(Y_n)_{n \geq 0}$ is an \defn{antitone strict supermartingale w.r.t.~$T$} if for all $n \geq 0$, we have $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\(
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
\).
%Any closed term for which an antitone ranking function exists is called \emph{antitone rankable}.
\end{definition}

@@

\begin{definition}\rm
Define an \defn{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{asparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{asparaenum}
\end{definition}

\begin{therm}[Sparse function]
%\begin{restatable}[Sparse function]{therm}{AntitonePartial}  
\label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
%\end{restatable}
\end{therm}

@@

Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs has not received much attention (with the exception of \citep{BeutnerO21}).
Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

@@

\begin{example}[Higher-order recursion]
\label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined by
\[
%\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} \, . \, 
\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
%\end{array}
\]
Let $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$. Then $\Xi\ F\ \underline x\ \underline y$ has the antitone sparse ranking function 
\[
F \, \underline{n_0} \, (F \, \underline{n_1} \, (\dots(\Xi\ F\, \underline x\, \underline n)\dots) \mapsto g_2(n),
\] 
where $g_2$ is as in Example~\ref{ex:unbiased random walk}, using the same antitone function too.

### 2021-10-16 LICS21/confluent-proofs.tex
for important technical words, changed \emph to \defn moved discussion of sum of powers from appendix to main body other minors

\paragraph{}
Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting. Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram). The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$. The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch. The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction). Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.

@@

\paragraph{}
If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, it will be shown that the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$. To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$. The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions. If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already. Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order). If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it. If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$. If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.

@@

\paragraph{}
If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
\end{proof}

@@

\paragraph{}
In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps. However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma~\ref{downAcrossUp}. To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s. In the case of Lemma~\ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s. In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term). The same argument that the process in Lemma~\ref{downAcrossUp} terminated is therefore applicable here too. At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
\end{proof}

@@

  \paragraph{}
  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order. It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$. Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$. Suppose further, for a contradiction, that $N_j$ is not a value. Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g. $\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$). In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
\end{proof}

@@

Example~\ref{ex:sum of powers}
Recall that
\begin{align*}
M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
\end{align*}

@@

It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.

>>>>

%\paragraph{}
Although the six cases in the definition of $\sim_c$ are defined separately, there are some commonalities worth noting. Each case (except a, which is symmetrical) has a right branch and a left branch (the branches of case b are unfortunately displayed the wrong way around in the diagram). The right branch has reduction steps at two positions, $\beta$ then $\alpha$, where $\beta > \alpha$ (i.e.~$\beta$ is inside $\alpha$), and in the left branch, the order of these reductions are swapped, so the reductions are at $\alpha$ then $\beta_1, \dots, \beta_n$. The position $\alpha$ of one of the reductions is unchanged, and the other reduction may have multiple (in cases e and f) or 0 (in cases b and e) images in the left branch. The images $(\beta_i)$ are all still inside (greater than or equal to) $\alpha$, and although they are not generally equal to $\beta$, the subskeletons at these positions have a similar shape to the subskeleton (initially) at position $\beta$, so that the reductions at these positions are the same type of reduction (e.g.~both $\beta$-reduction or both $\textsf{if}$-reduction). Case a is also similar, except that the relevant positions are all disjoint rather than contained in one another, and either branch could be considered the right branch.

@@

%\paragraph{}
If $Y \sim_c X \sim_c^* X_{cbv}$, and the reductions involved in $Y \sim_c X$ are not the last pair of reductions in $Y$ that aren't in CBV order, it will be shown that the order of the $\sim_c$s can be rearranged so that the sequence $Y \sim_c^* X_{cbv}$ is the canonical sequence given above, therefore $Y_{cbv} = X_{cbv}$. To be more precise about this rearrangement, consider the sequence of $\sim_c$ steps from $Y$ to $X_{cbv}$. The first step takes some sub-sequence of the reduction sequence $Y$ and swaps the order of those reductions. If $X = X_{cbv}$, then $X = Y_1 = Y_{cbv}$ already. Otherwise, consider the cases according to whether the image of this subsequence in $X$ occurs earlier than the next pair of reductions to be swapped (i.e.~the last pair of reductions in $X$ that occur not in CBV order). If not, then either the step $Y \sim_c X$ changes this subsequence into CBV order, or out of it. If it is into, then it was the last out-of-order pair in $Y$ therefore the whole sequence $Y \sim_c^* X_{cbv}$ is already in the canonical order and $Y_{cbv} = X_{cbv}$. If it switches the pair to the wrong order, then the result of that switch is the last out-of-order pair in $X$ therefore $Y = X_1$, and the sequence $X_1 \sim_c^* X_{cbv}$ is just a suffix of the sequence $X \sim_c^* X_{cbv}$ therefore $(X_1)_{cbv} = X_{cbv}$ and again $Y_{cbv} = X_{cbv}$.

@@

%\paragraph{}
If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ because the sequence $(Y_i)_i$ terminates immediately therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
\end{proof}

@@

%\paragraph{}
In the case that $(M,s) \Rightarrow^* (M_1,s_1)$ or $(M,s) \Rightarrow^* (M_2,s_2)$ by multiple steps, it is possible to repeatedly replace a pair of the form $A \Leftarrow B \Rightarrow C$ by $A \Rightarrow^* D \Leftarrow^* C$ by the construction above, but it is not immediate that this process terminates, because each $\Rightarrow$ or $\Leftarrow$ may be replaced by multiple, so the sequence of $\Rightarrow$s and $\Leftarrow$s from $(M_1,s_1)$ to $(M_2,s_2)$ may get longer at some steps. However, termination can be proved by noting that the structure of this process is identical to the process of swapping $\sim_\uparrow$ and $\sim_\downarrow$ steps in Lemma~\ref{downAcrossUp}. To be more precise, in this case there is a sequence of $\Leftarrow$ and $\Rightarrow$ steps, each of which has an associated reduction position and initial and final terms, and if a $\Leftarrow$ immediately precedes a $\Rightarrow$, they may be swapped to produce some number of $\Rightarrow$s, followed by some number of $\Leftarrow$s. In the case of Lemma~\ref{downAcrossUp}, there is a sequence of $\sim_\uparrow$ and $\sim_\downarrow$ steps, each of which has an associated reduction position and initial and final skeletons, and if a $\sim_\uparrow$ immediately precedes a $\sim_\downarrow$, they may be swapped to produce some number of $\sim_\downarrow$s followed by some number of $\sim_\uparrow$s. In both cases, the number and reduction positions of the resultant steps are determined by the case of $\sim_c$ that matches the way the initial reduction positions overlap, and the initial skeleton (or the skeleton of the initial term). The same argument that the process in Lemma~\ref{downAcrossUp} terminated is therefore applicable here too. At every stage, the sequence of reductions that results from one of the initial $\Rightarrow$s is a parallel sequence, and swapping a parallel sequence of $\Rightarrow$s with a $\Leftarrow$ always terminates, therefore each $\Rightarrow$ in turn can be moved past all of the $\Leftarrow$s, and the process as a whole will terminate in a state where all of the $\Rightarrow$s precede all of the $\Leftarrow$s, i.e.~a pair of reduction sequences $(M_1,s_1) \Rightarrow^* (M',s') \Leftarrow^* (M_2,s_2)$.
\end{proof}

@@

  %\paragraph{}
  There is therefore a reduction sequence $(M,s) = (N_0,s_0') \Rightarrow \dots \Rightarrow (N_m, s_m')$ which ends in a value ($N_m = M_n$) and is in CBV order. It doesn't immediately follow that this is the $\Rightarrow_{\cbv}$ reduction sequence starting from $(M,s)$ because the fact that $(N_i)_i$ is in CBV order only means that the reductions that do take place in this reduction sequence are in the correct order, not that the reduction sequence finishes once it reaches a value, or that there are no other reductions that would come earlier in CBV order that don't occur at all in $(N_i)_i$. Suppose that at some point $j$ in the reduction sequence, the reduction that occurs is not the next reduction in CBV order, but something else, and that the reduction that should be next is at position $\alpha = \cbv(N_i)$, and the position of the actual reduction $N_j \to N_{j+1}$ is $\beta$. Suppose further, for a contradiction, that $N_j$ is not a value. Either $\alpha = \cdot$, the root position, or the reduction at the root position is dependent in some way on the reduction at $\alpha$ (e.g. $\alpha = \textsf{if}_1$, or $\alpha = \underline{+};@_1$). In any case, all the reductions in the rest of the reduction sequence are after $\alpha$ in CBV order, therefore none of them are $\leq \alpha$, and there continues to be an unreduced redex at $\alpha$, therefore the reduction at the root position never occurs, therefore $N_m$ is not a value, which is a contradiction, therefore at the point (if any) where $(N_i)_i$ departs from the $\Rightarrow_{\cbv}$ reduction sequence starting at $(M,s)$, it has already reached a value, therefore $(M,s) \Rightarrow_{\cbv} \dots$ does reach a value eventually, i.e.~$M$ terminates with the reduction strategy $\cbv$ and samples $s$.
\end{proof}

@@

\iffalse
Example~\ref{ex:sum of powers}
Recall that
\begin{align*}
M =&\ (\lambda n. P[n]) \lfloor \tsample^{-1/2} \rfloor \\
P[n] =&\ (\lambda p. \Xi[p] n) (\lambda x. \Theta[x] n)) \\
\Xi[p] =&\ \tY \lambda f n. \tif{n = 0}{\underline 0}{p n + f (n-1)} \\
\Theta[x] =&\ \tY \lambda f n. \tif{n = 0}{\underline 1}{x \times f (n-1)}.
\end{align*}

@@

It wouldn't even be possible in this case to give a sparser version of the ranking function defined only at $M$ and $P[\underline n]$, because the amount that an antitone sparse ranking function must decrease at each $\tY$-reduction step depends on the value of the ranking function at the next term where it is defined, therefore it is necessary to have these intermediate steps in order for the ranking function to be able to change sufficiently slowly.
\fi
